Dissecting the relationship between protein structure and sequence variation

### 2015-05-20 manuscript/voronoi/mnv.tex
MN update + new figures and tables on best definition of WCN

    A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties. Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2014, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, meyer_geometric_2015}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{wilke_predicting_2005, echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based on elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}. \\

@@

    \begin{equation}
        \label{eqn:wcn_pwrl}
        \text{WCN}_{i} = \sum^N_{j\neq i} \frac{1}{{r_{ij}}^{\alpha=2}}.
    \end{equation}

@@

    where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment. We use DSSP software \citep[][]{kabsch_dictionary_1983} for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of \citet{tien_maximum_2013} to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins.

@@

    As a measure of local flexibility or fluctuation in different parts of the protein structure, the temperature factor ({\it B factor}) for all atoms in PDB files were extracted. Although, B factor is an atomic measure of flexibility and fluctuation in proteins, the backbone C$_\alpha$ B factor has become a very popular proxy measure of amino acid flexibility in the studies of protein dynamics and benchmarking of different Elastic Network Models of proteins. Alternatively, the site-specific fluctuation could be calculated from MD simulations. This was however impossible for this study for the large dataset of $209$ proteins considered in this work. \\

@@

    In order to determine the extent to which long-range amino acid interactions influence the general patterns of sequence variability and local flexibility measures, first we investigate the behavior of Contact Number and the Weighted Contact Number in predicting site-specific evolutionary rates (r4sJC) and the average B factor of the amino acid side-chains, for a wide range of the free parameters of the two packing density measures (i.e., $r_0\in[0\AA,50\AA]$ \& $\alpha\in[-30\AA,30\AA]$ as in Eqns. \ref{eqn:cn} \& \ref{eqn:wcn_pwrl}).

    The results for the dataset of $209$ monomeric enzymes are plotted in Figures \ref{fig:cn} \& \ref{fig:wcn}. The values of the free parameters of CN and WCN that result in the strongest correlations are tabulated in Table \ref{tab:cnwcn}. Evidently, the best-performing free parameters of both model indicate the non-negligible contributions of long-range amino acid interactions in the strengths of the observed correlations. The extent and significance of the influences however, cannot be deciphered solely based on the two common definitions of packing density: CN \& WCN. In the following section we present a methodology to segregate and quantify the role of local packing density from long-range interactions in on site-specific sequence variability and flexibility.

@@

    \subsection*{Voronoi Cell Volume as Proxy Measure of Local Packing Density and Flexibility in Proteins}

@@

    In order to assess the prediction power of site-specific variables derived from Voronoi tessellation, first the geometric centers of all side-chains for each of the proteins in dataset were calculated and used as the seeds of Voronoi polyhedra. We show in Appendix \ref{app:xx} that the choice of the geometric center of side chains as Voronoi seeds -- in contrast to other sets of atomic coordinates representative of individual sites in protein -- results in the strongest correlations of Voronoi cell properties with other site-specific structural and evolutionary characteristics of proteins.  \\

@@

    Depending on the choice of the Cartesian coordinates used, there exist degeneracy in the definition of some site-specific structural variables. For example, the quantity WCN is generally calculated from the coordinates of $C_\alpha$ atoms in the 3-dimensional structure of protein. The choice of $C_\alpha$ coordinates is however mainly driven by convenience in WCN calculation and there is no reason to believe this set of atomic coordinates is the best representative of individual sites in proteins. Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein \cite{soyer_voronoi_2000}. More recently, Marcos \& Echave (2014)\cite{marcos_too_2014} have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures. \\

@@

    \begin{figure}
        \begin{center}
        \begin{tabular}{cc}
            \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnhSC_r4sJC.pdf} & \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnhSC_bfSC.pdf}
        \end{tabular}
        \end{center}
        \caption{}
        \label{fig:wcnh}
    \end{figure}

    \begin{figure}
        \begin{center}
        \begin{tabular}{cc}
            \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_r4sJC.pdf} & \includegraphics[width=3.5in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_bfSC.pdf}
        \end{tabular}
        \end{center}
        \caption{Average absolute Spearman's correlation strengths of Weighted Contact Number (WCN$_\operatorname{P}$) with power-law kernel as given by Eqn. \ref{eqn:wcn_pwrl}, with site-specific evolutionary rates ({\it left plot}) and average side-chain B factor ({\it right plot}), for different values of the free parameter of the kernel $\alpha$. The solid black line represents the mean correlation strength in the entire dataset of 209 proteins at each value of $\alpha$, and the dashed black line indicates the median of the distribution. The green-shaded region together with the two red-dashed lines represent the $25\%$ \& $75\%$ quartiles of the correlation strength distribution.  Note that for $\alpha>0$ the sign of the correlation strength $\rho$ is the opposite of the sign of $\rho$ for $\alpha<0$. In addition $\rho$ is undefined at $\alpha=0$ and not shown in this plot. The parameter values at which the Spearman's correlation coefficient reaches the maximum over the entire dataset are given in Table \ref{tab:best_params}.}
        \label{fig:wcnp}
    \end{figure}

@@

>>>>

    A variety of site-specific structural characteristics have been proposed over the past decade to predict protein sequence evolution from structural properties. Among the most important and widely discussed are the Relative Solvent Accessibility (RSA) \citep[e.g.,][]{goldman_assessing_1998, bustamante_solvent_2000, conant_solvent_2009, franzosa_structural_2009, ramsey_relationship_2011, scherrer_modeling_2012, meyer_integrating_2013, meyer_cross-species_2013, yeh_local_2014, yeh_site-specific_2014, shahmoradi_predicting_2014, sikosek_biophysics_2014, meyer_geometric_2015}, Contact Number \citep[e.g.,][]{rodionov_sequence_1998, hamelryck_amino_2005, liao_protein_2005, bloom_structural_2006, huang_mechanistic_2014, marcos_too_2015, yeh_site-specific_2014, yeh_local_2014, shahmoradi_predicting_2014, meyer_geometric_2015}, measures of thermodynamic stability changes due to mutations at individual sites in proteins \citep[e.g.,][]{wilke_predicting_2005, echave_relationship_2014}, and measures of local flexibility, such as the Debye-Waller factor (hereafter B factor) \citep[e.g.,][]{liao_protein_2005, shih_evolutionary_2012, shahmoradi_predicting_2014} or flexibility measures based on elastic network models \citep[e.g.,][]{liu_sequence_2012} and Molecular Dynamics (MD) simulations \citep[e.g.,][]{shahmoradi_predicting_2014}. \\

@@

    \begin{equation}
        \label{eqn:wcn_pwrl}
        \text{WCN}_{i} = \sum^N_{j\neq i} {r_{ij}^{\alpha=-2}}.
    \end{equation}

@@

    where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment. We use DSSP software \citep[][]{kabsch_dictionary_1983} for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of \citet{tien_maximum_2013} to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins. \\

@@

    As a measure of local flexibility or fluctuation in different parts of the protein structure, the temperature factor ({\it B factor}) for all atoms in PDB files were extracted. Alternatively, the site-specific fluctuation could be calculated from MD simulations. This was however impossible for this study for the large dataset of $209$ proteins considered in this work. Although, B factor is an atomic measure of flexibility and fluctuation in proteins, the backbone C$_\alpha$ B factor has become a very popular proxy measure of amino acid flexibility in the studies of protein dynamics and benchmarking of different Elastic Network Models of proteins. Here, we use instead the average over all B factors of the side chain atoms if individual amino acids. We show in Appendix \ref{app:best_bf} that the average side chain B factor is the best representation of site-specific flexibility among all possible atomic B factors in an amino acid. \\

@@

    In order to determine the extent to which long-range amino acid interactions influence the general patterns of sequence variability and local flexibility measures, first we investigate the behavior of Contact Number and the Weighted Contact Number in predicting site-specific evolutionary rates (r4sJC) and the average B factor of the amino acid side-chains, for a wide range of the free parameters of the two packing density measures (i.e., $r_0\in[0\AA,50\AA]$ \& $\alpha\in[-30\AA,30\AA]$ as in Eqns. \ref{eqn:cn} \& \ref{eqn:wcn_pwrl}). \\

    The results for the dataset of $209$ monomeric enzymes are plotted in Figures \ref{fig:cn} \& \ref{fig:wcnp}. The values of the free parameters of CN and WCN that result in the strongest correlations are tabulated in Table \ref{tab:cnwcn}. Evidently, the best-performing free parameters of both model indicate the non-negligible contributions of long-range amino acid interactions in the strengths of the observed correlations. The extent and significance of the influences however, cannot be deciphered solely based on the two common definitions of packing density: CN \& WCN. In the following section we present a methodology to segregate and quantify the role of local packing density from long-range interactions in on site-specific sequence variability and flexibility.

@@

    \subsection*{Voronoi Cell Volume as a Proxy Measure of Local Packing Density and Flexibility in Proteins}

@@

    In order to assess the prediction power of site-specific variables derived from Voronoi tessellation, first the geometric centers of all side-chains for each of the proteins in dataset were calculated and used as the seeds of Voronoi polyhedra. We show in Appendix \ref{app:best_voronoi} that the choice of the geometric center of side chains as Voronoi seeds -- in contrast to other sets of atomic coordinates representative of individual sites in protein -- results in the strongest correlations of Voronoi cell properties with other site-specific structural and evolutionary characteristics of proteins.  \\

@@

    Depending on the choice of the Cartesian coordinates used, there exist degeneracy in the definition of some site-specific structural variables. For example, the quantity WCN is generally calculated from the coordinates of $C_\alpha$ atoms in the 3-dimensional structure of protein. The choice of $C_\alpha$ coordinates is however mainly driven by convenience in WCN calculation and there is no reason to believe this set of atomic coordinates is the best representative of individual sites in proteins. Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein \cite{soyer_voronoi_2000}. More recently, Marcos \& Echave (2015) \cite{marcos_too_2015} have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures. \\

@@

    \begin{figure}
        \begin{center}
        \begin{tabular}{cc}
            \includegraphics[width=3.3in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnSC_r4sJC.pdf} & \includegraphics[width=3.3in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_cnSC_bfSC.pdf}
        \end{tabular}
        \end{center}
        \caption{Average absolute Spearman's correlation strengths of Contact Number (CN) -- as defined by Eqn. \ref{eqn:wcnh} using Heaviside kernel -- with site-specific evolutionary rates ({\it left plot}) and average side-chain B factor ({\it right plot}), for different values of the free parameter of the kernel $r_0$. The solid black line represents the mean correlation strength in the entire dataset of 209 proteins at each value of $r_0$, and the dashed black line indicates the median of the distribution. The green-shaded region together with the two red-dashed lines represent the $25\%$ \& $75\%$ quartiles of the correlation strength distribution.  The parameter values at which the Spearman's correlation coefficient reaches the maximum over the entire dataset are given in Table \ref{tab:best_params}.}
        \label{fig:cn}
    \end{figure}

    \begin{figure}
        \begin{center}
        \begin{tabular}{cc}
            \includegraphics[width=3.3in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_r4sJC.pdf} & \includegraphics[width=3.3in]{../../wcn_best_definition/analysis/figures/get_quantiles/screen_plots/spcor_wcnpSC_bfSC.pdf}
        \end{tabular}
        \end{center}
        \caption{Average absolute Spearman's correlation strengths of Weighted Contact Number (WCN) -- as defined by Eqn. \ref{eqn:wcn_pwrl} using a power-law kernel -- with site-specific evolutionary rates ({\it left plot}) and average side-chain B factor ({\it right plot}), for different values of the free parameter of the kernel $\alpha$. The solid black line represents the mean correlation strength in the entire dataset of 209 proteins at each value of $\alpha$, and the dashed black line indicates the median of the distribution. The green-shaded region together with the two red-dashed lines represent the $25\%$ \& $75\%$ quartiles of the correlation strength distribution.  Note that for $\alpha>0$ the sign of the correlation strength $\rho$ is the opposite of the sign of $\rho$ for $\alpha<0$. In addition $\rho$ is undefined at $\alpha=0$ and not shown in this plot. The parameter values at which the Spearman's correlation coefficient reaches the maximum over the entire dataset are given in Table \ref{tab:xx}.}
        \label{fig:wcnp}
    \end{figure}

@@

% Tables

    \begin{table}[htbp]
    \caption{Best free parameter values of the Contact Number (CN) and the Weighted Contact Number with power-law kernel (WCN) that result in the strongest median Spearman's correlation ($\rho$) of WCN with four site-specific quantities (average side-chain B factor, evolutionary rates (r4sJC), sequence entropy, and \ddg rate) for the entire dataset of 209 proteins. The corresponding median correlation coefficients ($\rho$) are reported  inside parenthesis next to each parameter value in the table. \label{tab:best_params}}
    \bigskip
    \centerline{
    \begin{tabular}{|l|c|c|c|c|}
      \hline
      \multirow{3}{*}{Correlation} & \multicolumn{4}{|c|}{Best Performing Free Parameter of the Kernel} \\ \cline{2-5}
                        & Power-law & Exponential & Gaussian & Cutoff \\
                        & $\alpha$ & $\lambda~[\AA]$ & $\sigma~[\AA]$ & $r~[\AA]$ \\ \hline
      WCN--B factor     & -2.2 ($0.70$) & 3.8 ($0.70$) & 9.6 ($0.70$) & 13.0 ($0.69$) \\
      WCN--r4sJC        & -2.3 ($0.64$) & 3.4 ($0.64$) & 9.8 ($0.63$) & 14.8 ($0.62$) \\
      WCN--Seq. Entropy & -2.2 ($0.54$) & 3.4 ($0.53$) & 8.0 ($0.53$) & 15.2 ($0.52$) \\
      WCN--\ddg Rate    & -3.4 ($0.76$) & 2.0 ($0.78$) & 5.4 ($0.79$) & 9.8 ($0.75$) \\
      \hline
    \end{tabular}}
    \end{table}
